Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-10-20 crowdsource/article.tex
Description of the full method (PG's version)

A partial correlation approach \cite{sutera2014simple}  is a simple yet
effective solution to the problem of connectome inference from calcium imaging
data. This partial correlation statistics, equivalent to the inverse of the correlation
measure, has  the property to filter out indirect associations between neurons and
only gives a high degree of association between directly connected neurons. The
winning solution of the Connectomics challenge combines two main steps:
(i) A pre-processing of the data using signal processing filters; and (ii)
The inference of the network through partial correlation. The pre-processing procedure
highly improves the infered network quality by filtering out noise and taking into account
specifities of the fluorecense signals.

@@

The inference problem reduced to its simplest shape is the identification of
neuron pairs that are directly connected in the network. Intuitively, such
pairs should be related somehow and in particular their activities have to be
similar since their are dependent on the connectivity network. Therefore a way
to identify them is to measure, or at least estimate, how closely related two
neurons are regarding their activities. The most common measure for that
purpose is the (Pearson) correlation.  However, correlation also measures the
degree of association between neurons that are connected through others and so
connected indirectly. It means all related neurons are found but direct pairs (i.e., without any intermediates)
are indistinguishable from indirect ones.

A variant of this classic approach is the partial correlation
\citep{sutera2014simple}, also known as the inverse of the correlation, which
measures the degree of association between two neurons taking into account the
state of all other neurons. Under technical assumptions, the inverse
correlation can be interpreted as a measure of association between directly
connected pairs of neurons while filtering out indirectly connected pairs of
neurons. In genetics network, partial correlation have already shown to be
effective way to infer gene regulatory networks
\citep{de2004discovery,Schafer:2005}. This approach is fully described in
\cite{sutera2014simple}.

In the following paragraphs, we first describe the pre-processing of the signal,
then explained how to compute the partial correlation statistics.

Under the simplifying assumption that neurons are on-off units, characterised
by short periods of intense activity, or peaks, and longer periods of
inactivity, the first part of our algorithm consists in cleaning the raw
fluorescence data. More specifically, time-series are pre-processed using
standard signal processing filters in order to : (i) remove noise mainly due
to fluctuations independent of calcium, calcium fluctuations independent of
spiking activity, calcium fluctuations in nearby tissues that have been
mistakenly captured, or simply by the imaging process ; (ii) to account for
fluorescence low decay ; and (iii) to reduce the importance of high global
activity in the network. The first step of our pipeline is to smooth the
fluorescence signal by applying a low-pass filter for filtering out high
frequency noise:  
\begin{align} f_1(x^t_i) &= x^{t-1}_i + x^t_i + x^{t+1}_i,
\label{eq:symetric-median} \\ f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i +
0.8 x^{t-1}_i + x_i^t. \label{eq:weighted- asymetric-median} 
\end{align}

To have a signal that only has high magnitude around instances where the
spikes occur, the second step of our pipeline transforms the time-series into
its backward difference 
\begin{align} g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i.
\label{eq:high-pass-filter}
\end{align}

To filter out small variations in the signal obtained after applying the
function $g$, as well as to eliminate negative values, we use the following
hard-threshold filter
\begin{align}\label{eqn:hfilter}
h(x^{t}_i) &= x^{t}_i \mathbb{1}(x^{t}_i \geq \tau) \text{ with } \tau > 0,
\end{align}
where $\tau$ is the threshold parameter and $\mathbb{1}$ is the indicator
function.

The objective of the last step of our filtering procedure is to decrease the
importance of spikes that occur when there is high global activity in the
network with respect to spikes that occur during normal activity. Indeed, we
have conjectured that when a large part of the network is firing, the rate at
which observations are made is not high enough to be able to detect
interactions, and that it would therefore be preferable to lower their
importance by changing their magnitude appropriately. Additionally, it is well-known
that neurons may also spike because of a high global activity
\citep{stetter2012model}. In such context, detecting pairwise neuron
interactions from the firing activity is meaningless. As such, the signal
output by h is finally applied to the following function
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}}, \label{eq:magnify-filter}
\end{align}
whose effect is to magnify the importance of spikes that occur in cases of low
global activity (measured by $\sum_j x^t_j$).

Under the assumption that fluorescence level neuron are jointly Gaussian,
partial correlation thus measures conditional dependencies between
variables ; therefore it should naturally only detect direct associations
between neurons and filter out spurious indirect effects as mentioned before.

Practically speaking, the computation of all $p_{i,j}$ coefficients requires
the estimation of the covariance matrix $\Sigma$ and then computing its
inverse. Given that typically we have more samples than neurons, the covariance
matrix can be inverted in a straightforward way. We nevertheless obtained some
improvement by replacing the exact inverse with an approximation using only the
$M$ first principal components \citep{bishop2006pattern} (more details in
\citep{sutera2014simple}).

\begin{algorithm}[ht]
\begin{algorithmic}
 \algsetup{linenodelimiter= }

\STATE Here the pseudo-code.
% \STATE \textbf{Given} the sampled observation of $p$ neurons for $T$ time intervals.\\[2ex]

% \FOR{$(f,\tau) \in (\{f_1, f_2\}, \{\tau_1, \tau_2, \dots \})$}

% \STATE \vspace{1em} \textbf{Phase 1:} Apply the composed function $w \circ h \circ g \circ f$ to each fluorescent signal.
% \FOR{$x \in \{1,\dots,T\} $}
% \STATE $\bar{x}^t_i  \leftarrow w \circ h_{\tau} \circ g \circ f (x^t_i)$
% \ENDFOR\\[2ex]

% \STATE \textbf{Phase 2:} Compute \textit{Partial correlation} coefficient $p_{i,j}$ for each neuron pair $(i,j)$.
% \FOR{all pair $(i,j)$}
% \STATE Estimation of the covariance matrix $\Sigma$
% \STATE Computing $\Sigma^{-1}$, the inverse of $\Sigma$
% \STATE $p_{i,j} =
% -\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}$

% \ENDFOR\\[2ex]
% \ENDFOR

\end{algorithmic}
\caption{Inverse covariance pseudo-code \textcolor{red}{TO DO}}
\label{alg:inverse-covariance-pseudocode}
\end{algorithm}

% \begin{algorithm}[h]
%  \KwData{Given the sampled observation $\{ x^t_i \in \mathbb{R} | i \in V, t = 1, \dots, T \}$ of $p$ neurons for T time intervals}
%  % \KwResult{how to write algorithm with \LaTeX2e }

%  \textbf{Phase 1:} Apply the composed function $w \circ h \circ g \circ f$ to each fluorescent signal.\;\;
% \ \ For

% {\Indp
%  \For{$x \in \{1,\dots,T\} $}{
% 		$\bar{x}^t_i  \leftarrow w \circ h \circ g \circ f (x^t_i)$
%  }
%  }

%  \While{not at end of this document}{
%   read current\;
%   \eIf{understand}{
%    go to next section\;
%    current section becomes this one\;
%    }{
%    go back to the beginning of current section\;
%   }
%  }
%  \caption{Inverse covariance pseudo-code}
%  \label{alg:inverse-covariance-pseudocode}
% \end{algorithm}

>>>>

The winning solution of the Connectomics challenge \citep{sutera2014simple}
combines two main steps: (i) a pre-processing of the raw signals to detect
neural peak activities; and (ii) the inference of the network through partial
correlation statistics. The pre-processing step is crucial to obtain optimal
performance. It combines several standard filtering functions with three main
goals: to remove the noise introduced at different levels, to account for
fluorescence low decay, and to reduce the importance of high global activity in
the network. From the filtered signals, the strength of the connection between two
neurons is then assessed using partial correlations. The partial correlation
between two random variables measures their conditional dependencies, i.e.,
their associations when the effect of all the other variables has been
removed. With respect to plain correlation, the main advantage of partial
correlation for network inference is to only give a high degree of association
to directly connected neurons and thus to filter out spurious indirect
associations. Practically, partial correlations are also easily computed for
all pairs of neurons at once from a simple inversion of the data covariance
matrix.

@@

Our approach consists of two main steps: (i) a pre-processing of the raw
signals with several signal processing filters; and (ii) the inference of the
network through partial correlation statistics. We describe the details of
these two steps below as well as two techniques that slightly improve
performance, averaging and edge orientation. Further motivations for all steps
are given in \citep{sutera2014simple}. A simplified method is also proposed in
the same paper with only slightly lower performance.

\paragraph{Pre-processing.}

Under the simplifying assumption that neurons are on-off units, characterised
by short periods of intense activity, or peaks, and longer periods of
inactivity, the first part of our algorithm consists in cleaning the raw
fluorescence data. This pre-processing applies successively four
filtering functions to the original signal:
\begin{itemize}
  \item The first function smooths the fluorescence signal, using one of the following
low-pass filters for filtering out high frequency noise:
\begin{align} f_1(x^t_i) &= x^{t-1}_i + x^t_i + x^{t+1}_i, \label{eq:symetric-median} \\
  f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_i^t.\label{eq:weighted-asymetric-median}\\
  f_3(x^t_i) &= x^{t-1}_i + x^{t}_i + x^{t+1}_i + x^{t+2}_i, \label{eq:asymetric-median-forward} \\
f_4(x^t_i) &=  x_i^t + x^{t+1}_i  + x^{t+2}_i + x^{t+3}_i. \label{eq:asymetric-median}
\end{align}
\item To have a signal that only has high magnitude around instances where the
spikes occur, the second filter function transforms the time-series into
its backward difference 
\begin{align} g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i.
\label{eq:high-pass-filter}
\end{align}
\item To filter out small variations in the signal obtained after applying the
function $g$, as well as to eliminate negative values, we then use the following
hard-threshold filter
\begin{align}\label{eqn:hfilter}
h(x^{t}_i) &= (x^{t}_i)^c \mathbb{1}(x^{t}_i \geq \tau) \text{ with } \tau > 0,
\end{align}
where $\tau$ is the threshold parameter, $c$ is equal to 0.9, and $\mathbb{1}$ is the indicator
function.
\item The last filtering functions is defined as:
\begin{align}
 w^*(x^{t}_i) &= {(x^{t}_i + 1 )^{\left (1 + \frac{1}{\sum_{j} x^{t}_j}\right )}}^{k(\sum_{j} x^{t}_j)}
\end{align}
where the function $k$ is a piecewise linear function optimised
separately for each filter $f_1$, $f_2$, $f_3$ and $f_4$ (see the
implementation for full details). The effect of this filtering is to
magnify the importance of spikes that occur in cases of low global
activity (measured by $\sum_j x^t_j$).
\end{itemize}
The pre-processed times-series are then obtained by the application of
the following function: $w^*\circ h \circ g \circ f_i$ (with
$i=1$, 2, 3, or 4).

\paragraph{Partial correlation.}
In the second main step of our approach, partial correlation is used as a
measure of the strength of the connection between neurons. Considering each
time point of the pre-processed signals as an independent measurement
of $p$ random variables, one for each of the $p$ neurons,
the partial correlation coefficient $p_{i,j}$ between neurons $i$ and $j$ is
defined by:
\begin{equation}
p_{i,j} =
-\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}, \label{eq:inverse}
\end{equation}
where $\Sigma^{-1}$, known as the precision or concentration matrix, is the
inverse of the covariance matrix $\Sigma$ of the data. The computation of all
$p_{i,j}$ coefficients thus requires the estimation of the covariance matrix
$\Sigma$ and then computing its inverse. Given that typically we have more
samples (i.e., time points) than neurons, the covariance matrix can be inverted
in a straightforward way. We nevertheless obtained some improvement by
replacing the exact inverse with an approximation using only the $M$ first
principal components, with $M=0.8p$ \citep{bishop2006pattern}.

\paragraph{Averaging.}
The performance of our method is sensitive to the values of the pre-processing
parameters: the parameter $\tau$ of the hard-threshold filter $h$ (see Equation
\ref{eqn:hfilter}), and the choice of the low-pass filter (among $\{f_1,
f_2, f_3, f_4\}$). To improve robustness, we average partial correlation
statistics obtained for all pairs $(\tau,\mbox{low-pass filter}) \in
\{0.100,0.101,\ldots,0.209\}\times \{f_1, f_2, f_3, f_4\}$. In this averaging,
functions $f_1$, $f_2$, $f_3$, and $f_4$ were given weights respectively equal
to $0.383$, $0.345$, $0.004$, and $0.268$.

\paragraph{Edge orientation.}

Partial correlation is a symmetric measure that can not be used to predict edge
orientation. A post-processing heuristic was proposed to derive an asymmetric
association score from the partial correlation coefficient. This heuristic is
based on the observation that if there is directed edges from $i$ to $j$ then
the activation of $i$ should be followed by the activation of $j$ after a
slight delay. For every pairs $(i,j)$, we computed the following term:
\begin{align}
s_{i,j} = \sum_{t=1}^{T - 1} \mathbb{1}((x_j^{t+1} - x_i^t) \in \left[\phi_1, \phi_2\right])
\end{align}
that could be interpreted as an image of the number of times that neuron $i$
activates neuron $j$. $\phi_1$ and $\phi_2$ are parameters whose values have
been chosen equal to $0.2$ and $0.5$, respectively. The
final association score for an edge from $i$ to $j$ was
then the following:
\begin{align}
q_{i,j} = weight * p_{i,j} + (1-weight) * (s_{i,j}-s_{j,i}),
\label{eqn:qij}
\end{align}
with $p_{i,j}$ the average partial correlation coefficient and the parameter
$weight$ chosen close to 1 ($weight=0.997$). The second term is expected to be
greater (resp. lower) than 0 when $i$ more often activates $j$ than the
opposite and therefore in this case it makes $q_{i,j}$ greater (resp. lower)
than $q_{j,i}$.
